---
title: "County Data"
author: "Leona Odole"
date: "2025-04-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(R.matlab)
library(readxl)
library(reshape2)
```

So first I want to have the data be clean and in a workable format

```{r}
countydataraw <- readMat("./data/CountyMatrix_raw.mat")
countydata <- as.data.frame(countydataraw$d)
countydata <-`colnames<-`(countydata, colnames)
countydata$FIPS <- countydataraw$fips
```

Matlab stores NA values as NANs so I want to convert them 

```{r}
# since the dataset is small I will just go through it one by one even though its inefficient

# this funciton reassigns NaNs to NAs, could later repurpose to handle any replacement 
NaNtoNA <- function(dataset){
  copyds <- dataset
  for(row in 1:nrow(dataset)){
    for(col in 1:ncol(dataset)){
      if (is.nan(dataset[row,col])){
        copyds[row,col] <- NA
      }
    }
  }
  return(copyds)
}

```


```{r}
countydata <- NaNtoNA(countydata)
```

Just from inspecting the data it seems like most of the missing data is in the medicare reporting section. 

Question: For which diseases do we have a complete dataset? 

```{r}
names(which(colSums(is.na(countydata)) > 0))

```
So everything has missing data except for the fips column (the labels of the counties) some of it like lat long could be completed manually

Question: What percentage of the data is missing? 

```{r}
sum(is.na(countydata))/(nrow(countydata)*ncol(countydata))
```

approx 18 % of the data is missing 


Q: How are the different variables correlated with one another? 
```{r}
cormat <- round(cor(countydata, use = "pairwise.complete.obs"),2)
melted_cor <- melt(cormat)
ggplot(data = melted_cor, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile(color = "white")+
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
   name="Pearson\nCorrelation") +
  theme_minimal()+ 
 theme(axis.text.x = element_text(angle = 90, vjust = 1, 
    size = 12, hjust = 1))

```
There aren't really a lot of strong correlations outside the the racial/demographic factors I suspect because they are dependent on one another. 
Something else to note about the data is that the disease data is mostly given either in percentage (CDC data) or from the wonderdata i think per 1000? 
